Insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry

### 2019-04-08 analysis/paper/paper.Rmd
add abstract r code

---
title: "Insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Franziska Ruëff^2^ 
  - Claudia Pföhler^3^
  - Kathrin Scherer Hofmeier^4^
  - Margitta Worm^1^
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    #bookdown::html_document2:
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

# Abstract
Insect-venom elicited anaphylaxis is a common hypersensitivity reaction which may be life-threatening. Using the data from the European Anaphylaxis Registry (11596 cases in total) we identified insect-venom elicited anaphylaxis cases (n = 4482) and analyzed these in comparison to anaphylaxis elicited by other elicitors (n = 7114).

The data show that 68.57% of all insect elicited cases were elicited by yellow jackets, followed by bees (21.86%). The insect venom elicited cases occurred mostly in outdoor places (44.65%) patients' homes (12.87%) or urban places (9.616%).

Skin, gastrointestinal and respiratory symptoms occurred less frequently in insect elicited cases of anaphylaxis, whereas cardiologic symptoms (with hypotension, collapse, and loss of consciousness) were more frequent.  Intramuscular adrenaline (as a first-line therapy) was administered significantly less often in insect venom elicited cases (4.04%, p < 0.0001). The mortality rate in insect anaphylaxis was comparable (0.156%) to other cases (0.295%, p = 0.174).

@@

Insect venom elicited anaphylaxis in contrast to other elicitors showed a significant seasonal fluctuation and was most frequently reported from May till October. Their proportion to other cases during the summer seasons reached 60 % and was as under 1% of cases during winter. Yellow-jacket was the most prominent causing-insect reported in these cases foolowed by bee (Fig \@ref(fig:insectstime)).

```{r insectsgeo, fig.cap="Proportion of anaphylaxis cases elicited by specific insects in reporting countries. Less frequent insects were groupeped together as 'other'. Countries which reported less than 20 cases of anaphylaxis due to insect venom were not illustrated in this figure."}
plot.proportions(rdb, "d_centres_country","q_340_insects",5)+
  labs(x = "Country",y="Proportion of IVA cases",fill = "Species")
```

@@

```{r}
ana_outcome_matched <- function(data = data4,
         outcome = "d_severity_rmr",
         matching_var = "q_340_insects",
         confounders_cat = c("b_sex", "cc_cardio"),
         confounders_int = c("d_age"),
         selected_predictors = c(1:4)
         ){
  possible_outcomes <- c("d_severity_rm",
                         "d_severity_rmr",
                         "severity_brown",
                         "q_116_VAS_v7",
                         "ANAscore")
  predictors <- c(matching_var, confounders_cat,confounders_int)
  df <- data %>% # first get the data frame
   filter(!!as.name(matching_var) %in% c("yellow jacket", "bee")) %>% 
   select(!!!matching_var,
          !!!outcome,
          !!!confounders_cat,
          !!!confounders_int
          ) %>%
          {.[complete.cases(.),]}
  ### we need to match the cases now to the first level of theoutcome variable
  ref_level <- levels(df[,matching_var] %>% droplevels())
  matched <- df %>% 
    mutate(match = ifelse(!!as.name(matching_var)==ref_level[1],T,F),
           out=ifelse(!!as.name(outcome)=="severe",1,0)) %>% 
    matchit(formula = as.formula(
      paste("match ~ ",paste0(confounders_cat),"+", paste0(confounders_int,collapse = " + ")))) %>% 
    match.data() 
  ## IFELSE
  if(outcome %in% possible_outcomes[c(2,3)]){
    plot <- matched %>% 
      ggplot(aes(!!as.name(matching_var),
               fill = !!as.name(outcome)))+
      geom_bar()+
      facet_grid(reformulate(confounders_cat[1],confounders_cat[2]))

@@

temp <- lapply(1:5,do_cof_tab) %>% do.call(what = rbind) %>%
                          as_tibble() %>%
  {mutate(.,
         Estimate = as.numeric(Estimate),
         Pr...t.. = as.numeric(Pr...t..)
  )} %>%
          {cbind(.,outcome = lapply(possible_outcomes,
                          rep,
                          nrow(.)/5) %>% unlist)
          } %>% 
           filter(.rownames!="(Intercept)")
gridExtra::grid.arrange(output_severity[[2]]$plot,
                        output_severity[[3]]$plot,
                        output_severity[[1]]$plot,
                        output_severity[[4]]$plot,
                        output_severity[[5]]$plot,
                        temp %>% 
  ggplot(aes(.rownames,Pr...t.., color=outcome))+
  geom_segment(x=0.5,xend=4.5,y=log10(0.05),yend=log10(0.05))+
  scale_y_log10()+
  geom_jitter()+
  theme(axis.text.x = element_text(angle=40, hjust=1))
  )#+
  #geom_line(aes(group=outcome))
```

@@

```{r}
output_severity <- lapply(possible_outcomes,
                         function(x){
                         ana_outcome_matched(data = data4,
                            outcome = x,
                            matching_var = "q_340_insects",
                            confounders_cat = c("tryp_cat", "cc_cardio"),
                            confounders_int = c("d_age"),
                            selected_predictors = c(1,2,4))
                           }
)

@@

Based on the above we see that the lower tryptase significantly decreased the severity of

>>>>

---
title: "Insect venom anaphylaxis: a case - control study of the European Anaphylaxis Registry"
author:
  - Wojciech Francuzik^1^
  - Sabine Dölle-Bierke^1^
  - Franziska Ruëff^2^ 
  - Claudia Pföhler^3^
  - Kathrin Scherer Hofmeier^4^
  - Margitta Worm^1^
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    #bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

# Abstract
Insect-venom elicited anaphylaxis is a common hypersensitivity reaction which may be life-threatening. Using the data from the European Anaphylaxis Registry (`r nrow(data)` cases in total) we identified insect-venom elicited anaphylaxis cases (n = `r {data$d_elicitor_gr5=="insects"} %>% sum()`) and analyzed these in comparison to anaphylaxis elicited by other elicitors (n = `r {data$d_elicitor_gr5!="insects"} %>% sum()`).

The data show that `r {{rdb$q_340_insects=="yellow jacket"} %>% sum(na.rm=T)/sum({!is.na(rdb$q_340_insects)})*100} %>% {round(.,1)}`% of all insect elicited cases were elicited by yellow jackets, followed by bees (`r {{rdb$q_340_insects=="bee"} %>% sum(na.rm=T)/sum({!is.na(rdb$q_340_insects)})*100} %>% {round(.,1)}`%). The insect venom elicited cases occurred mostly in outdoor places (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(4),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%) patients' homes (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(8),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%) or urban places (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(q_152_location) %>% summarize(n=n()) %>% {.[c(5),2] %>% sum()/(.[,2] %>% sum)*100} %>% round(1)`%).

Skin, gastrointestinal and respiratory symptoms occurred less frequently in insect elicited cases of anaphylaxis, whereas cardiologic symptoms (with hypotension, collapse, and loss of consciousness) were more frequent.  Intramuscular adrenaline (as a first-line therapy) was administered significantly less often in insect venom elicited cases (`r rdb %>% filter(!is.na(q_340_insects)) %>% group_by(d_522_adren_agg) %>% summarize(n=n()) %>% {.[2,2]/(.[1,2])*100} %>% round(1)`% vs `r rdb %>% filter(is.na(q_340_insects)) %>% group_by(d_522_adren_agg) %>% summarize(n=n()) %>% {.[2,2]/(.[1,2])*100} %>% round(1)`, p < 0.0001). The mortality rate in insect anaphylaxis was comparable (`r `%) to other cases (0.295%, p = 0.174).

@@

Insect venom elicited anaphylaxis in contrast to other elicitors showed a significant seasonal fluctuation and was most frequently reported from May till October. Their proportion to other cases during the summer seasons reached 60% and was as under 1% of cases during winter. Yellow-jacket was the most prominent causing-insect reported in these cases foolowed by bee (Fig \@ref(fig:insectstime)).

```{r MORaltogether}

```{r insectsgeo, fig.cap="Proportion of anaphylaxis cases elicited by specific insects in reporting countries. Less frequent insects were groupeped together as 'other'. Countries which reported less than 20 cases of anaphylaxis due to insect venom were not illustrated in this figure."}
```

@@

```{r,fig.height=8}
require(broom)

ana_outcome_matched <- function(data = data4,
         outcome = "d_severity_rmr",
         matching_var = "q_340_insects",
         confounders_cat = c("b_sex", "cc_cardio"),
         confounders_int = c("d_age"),
         selected_predictors = c(1:4)
         ){
  possible_outcomes <- c("d_severity_rm",
                         "d_severity_rmr",
                         "severity_brown",
                         "q_116_VAS_v7",
                         "ANAscore")
  predictors <- c(matching_var, confounders_cat,confounders_int)
  df <- data %>% # first get the data frame
   filter(!!as.name(matching_var) %in% c("yellow jacket", "bee")) %>% 
   select(!!!matching_var,
          !!!outcome,
          !!!confounders_cat,
          !!!confounders_int
          ) %>%
          {.[complete.cases(.),]}
  ### we need to match the cases now to the first level of theoutcome variable
  ref_level <- levels(df[,matching_var] %>% droplevels())
  matched <- df %>% 
    mutate(match = ifelse(!!as.name(matching_var)==ref_level[1],T,F),
           out=ifelse(!!as.name(outcome)=="severe",1,0)) %>% 
    matchit(formula = as.formula(
      paste("match ~ ",paste0(confounders_cat),"+", paste0(confounders_int,collapse = " + ")))) %>% 
    match.data() 
  ## IFELSE
  if(outcome %in% possible_outcomes[c(2,3)]){
    plot <- matched %>% 
      ggplot(aes(!!as.name(matching_var),
               fill = !!as.name(outcome)))+
      geom_bar()+
      facet_grid(reformulate(confounders_cat[1],confounders_cat[2]))

@@

temp <- lapply(1:5,do_cof_tab) %>% do.call(what = rbind) %>%
                          as_tibble() %>%
  {mutate(.,
         Estimate = as.numeric(Estimate),
         Pr...t.. = as.numeric(Pr...t..)
  )} %>%
          {cbind(.,outcome = lapply(possible_outcomes,
                          rep,
                          nrow(.)/5) %>% unlist)
          } %>% 
           filter(.rownames!="(Intercept)")
gridExtra::grid.arrange(output_severity[[2]]$plot,
                        output_severity[[3]]$plot,
                        output_severity[[1]]$plot,
                        output_severity[[4]]$plot,
                        output_severity[[5]]$plot,
                        temp %>% 
  ggplot(aes(.rownames,Pr...t.., color=outcome))+
  geom_segment(x=0.5,xend=4.5,y=log10(0.05),yend=log10(0.05))+
  scale_y_log10()+
  geom_jitter()+
  theme(axis.text.x = element_text(angle=40, hjust=1))+
    annotate("text",x=4,y = c(log10(c(0.0001,0.00001,.000001,.0000001,.00000001))),
             label = c(1776,1780,1206,234,1780))
  )#+
  #geom_line(aes(group=outcome))

ggplot(temp,aes(.rownames,Pr...t.., color=outcome))+
  geom_segment(x=0.5,xend=4.5,y=log10(0.05),yend=log10(0.05))+
  scale_y_log10()+
  geom_jitter(width = 0.2)+
  theme(axis.text.x = element_text(angle=40, hjust=1))+
    annotate("text",x=4,y = c(0.0001,0.00001,.000001,.0000001,.00000001)*10,
             label = c("n=1776","n=1780","n=1206","n=234","n=1780"),size = 3)+
  labs(x = "",y="p value")
```

@@

```{r,fig.height=8}
output_severity <- lapply(possible_outcomes,
                         function(x){
                         ana_outcome_matched(data = data4,
                            outcome = x,
                            matching_var = "q_340_insects",
                            confounders_cat = c("tryp_cat", "cc_cardio"),
                            confounders_int = c("d_age"),
                            selected_predictors = c(1,2,4))
                           }
)

@@

Based on the above we see that the severity of anaphylaxis significantly depends on tryptase concentration (here under 8 µg/ml). Only in the VAS scale this did not sho as significant.
There might be problems with the vas scale...
